Supplemental Materials for the paper: \\ Pesticide exposure in small streams in Germany

### 2016-10-12 article.tex
reanalysed with only data for < 100km2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas they may be at high risk from pesticide pollution. 
However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales. 
We used data from German governmental water quality monitoring to quantify the drivers of pesticide risk and to assess pesticide risk in small streams on a large scale. 
The data set comprised of 1,766,104 measurements related to 24,743 samples in 2,301 sampling sites and of 478 pesticides.  %see do_overview.R
We investigated the influence of agricultural land use, catchment size, as well as precipitation and seasonal dynamics on pesticide risk using new statistical modeling techniques that explicitly consider the limit of quantification. 
Agriculture land use lead to an 3.4-fold increase in exceedance of risk thresholds once the proportion of agriculture in a catchment exceeded 25 percent.
Precipitation increased measured pesticide risk xxx-fold. 
The highest concentrations were found during summer.
%%%! Add these numbers also to results section!
Risk thresholds were exceeded in 27 percent of small streams, with neonicotinoid and other insecticides exhibiting highest frequencies of exceedance. 
We conclude that pesticides from agricultural land use are a major threat to small streams and their biodiversity, and that governmental pesticide sampling should be adapted to precipitation events in summer.

@@

@@

%% -------------------------------------------------------------------------
\section{Methods}
\subsection{Data compilation}
We requested pesticide monitoring data from sampling sites that can be classified as SWB (catchment sizes $\mathrm{< 100km^2}$ according to the WFD) from all 13 non-city federal states of Germany (see Supplemental Table~S1 for the abbreviations of federal state names). 
The request encompassed only the years 2005 to 2015.
We homogenized and unified all data provided by the federal states into a database and implemented a robust data cleaning workflow (see Supplemental Figure~S1 for details) \citep{poisot_best_2015}.

@@

We fitted this model separately to each compound with a RAC, measured in least 1000 samples and with more than 5\% of values above LOQ (n = 24 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the 24 modeled compounds we used a random effect meta-analysis for each model coefficient separately \citep{harrison_getting_2011}. 
The results of individual compounds are provided in the Supplemental Table~S4 and Figure~S7.
The meta-analysis was performed using the metafor package \citep{Viechtbauer_2010}.

@@

In total 478 different compounds used as pesticides and their metabolites were measured at least once (Supplemental Table~S2). 
Most of the compounds were herbicides (179), followed by insecticides (117) and fungicides (109). % see overview.R
Most samples were taken in the months April till October, with fewer samples during winter (see Supplemental Figure~S3).
Only 4\% (71,113) of all measurements were detects above the limit of quantification (LOQ).
We found substantial differences in the spectra of analyzed pesticides between federal states (Figure~\ref{fig:fig2}).
The number of different pesticides per state ranged from 57 (Saarland, SL) to 236 (Rhineland-Palatinate, RP) (Supplemental Table~S1).
Hierarchical clustering revealed that RP and NI analysed a distinct compound spectra compared to the cluster of other states.
Although, there was high variation within the remaining states these could not be separated (Figure~\ref{fig:fig2}, also Supplemental Figures~S4 and S5).
%%% unschöner satz: bearbeiten!

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure2.pdf}
  \caption{Compound spectra of the different federal states. Left: Barcode plot - each vertical line is an analysed compound. Right: MDS ordination. 
  Colors according to three groups determined by hierarchical clustering (see Supplemental Figure~S4).}
  \label{fig:fig2}
\end{figure}

%%%! Use a black only plot!?

@@

\subsection{Pesticides in small water bodies}
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure~\ref{fig:fig3}) and the low influence of catchment size on RAC exceedances (Figure~\ref{fig:fig4}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. 
This catchment size corresponds to a stream width of approximately 2~meters (see Supplemental Figure~S2).

The dataset comprised 12,710 samples from 1,295 SWB.
In 346 (26.7\%) of these streams RQ were exceeded.
In total 1,173 samples (0.3\% of all and 5.6\% of samples with detects) showed RQ higher than 1 were observed.
Neonicotinoid insecticides and Chlorpyrifos showed the highest RQ exceedances (Figure~\ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos the RAC was less than LOQ, therefore, all detections have a RQ~\textgreater~1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin also showed high exceedances of RQ (29.5, 14.2 and 21.4 \% of all samples with detects).
The highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Where analysed, metabolites exhibited the highest detection rates (for example, Metazachlor sulfonic acid was detected in 82\% of all samples were it was analysed, see also Supplemental Figure~S9).
Glyphosate was the compound with the highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure~\ref{fig:fig6}).
In 44.8\% of samples more than one compound was quantified, with a maximum of 54 different compounds in one sample (Supplemental Figure~S8).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas they may be at high risk from pesticide pollution. 
However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales. 
We used data from German governmental water quality monitoring to quantify the drivers of pesticide risk and to assess pesticide risk in small streams on a large scale. 
The data set comprised of 1,766,104 measurements related to 24,743 samples in 2,301 sampling sites and of 478 pesticides.  %see do_overview.R
We investigated the influence of agricultural land use, catchment size, as well as precipitation and seasonal dynamics on pesticide risk using new statistical modeling techniques that explicitly consider the limit of quantification. 
Agriculture land use lead to an 3.4-fold increase in exceedance of risk thresholds once the proportion of agriculture in a catchment exceeded 25 percent.
Precipitation increased measured pesticide risk xxx-fold. 
The highest concentrations were found during summer.
%%%! Add these numbers also to results section!
Risk thresholds were exceeded in 26 percent of small streams, with neonicotinoid and other insecticides exhibiting highest frequencies of exceedance. 
%%%! Add these numbers also to results section!
We conclude that pesticides from agricultural land use are a major threat to small streams and their biodiversity, and that governmental pesticide sampling should be adapted to precipitation events in summer.

@@

%%%! Check all numbers
%%%! Check all references to supplement

@@

%% -------------------------------------------------------------------------
\section{Methods}
\subsection{Data compilation}
We requested pesticide monitoring data from sampling sites that can be classified as SWB (catchment sizes $\mathrm{< 100km^2}$ according to the WFD) from all 13 non-city federal states of Germany (see Supplemental Table~S1 for the abbreviations of federal state names). 
%%%! Hier vielleicht noch auf ganz kleine Gewässer eingehen?
The request encompassed only the years 2005 to 2015.
We homogenized and unified all data provided by the federal states into a database and implemented a robust data cleaning workflow (see Supplemental Figure~S1 for details) \citep{poisot_best_2015}.

@@

We fitted this model separately to each compound with a RAC, measured in least 1000 samples and with more than 5\% of values above LOQ (n = 23 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the 24 modeled compounds we used a random effect meta-analysis for each model coefficient separately \citep{harrison_getting_2011}. 
The results of individual compounds are provided in the Supplemental Table~S4 and Figure~S7.
The meta-analysis was performed using the metafor package \citep{Viechtbauer_2010}.

@@

In total 478 different compounds used as pesticides and their metabolites were measured at least once (Supplemental Table~S2). 
Most of the compounds were herbicides (179), followed by insecticides (117) and fungicides (109). % see overview.R
Most samples were taken in the months April till October, with fewer samples during winter (see Supplemental Figure~S3).
Only 4\% (71,113) of all measurements were detects above the limit of quantification (LOQ).
We found substantial differences in the spectra of analyzed pesticides between federal states (Figure~\ref{fig:fig2}).
The number of different pesticides per state ranged from 57 (Saarland, SL) to 236 (Rhineland-Palatinate, RP) (Supplemental Table~S1).
Hierarchical clustering revealed that RP and NI analysed a distinct compound spectra compared to the cluster of other states.
Although, there was high variation within the remaining cluster this could not be further split (Figure~\ref{fig:fig2}, also Supplemental Figures~S4 and S5).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure2.pdf}
  \caption{Compound spectra of the different federal states. Left: Barcode plot - each vertical line is an analysed compound. Right: MDS ordination. 
  Colors according to two clusters determined by hierarchical clustering (see Supplemental Figure~S4).}
  \label{fig:fig2}
\end{figure}

@@

\subsection{Pesticide risk in small water bodies}
We found RAC exceedances in 26\% of sampling sites. 
Neonicotinoid insecticides and Chlorpyrifos showed the highest RQ (Figure~\ref{fig:fig6}).
For Thiacloprid and Chlorpyrifos the RAC was less than LOQ, therefore, all detections have a RQ~\textgreater~1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin also showed high exceedances of RQ (26.7, 14.1 and 21.1 \% of all samples~\textgreater~LOQ).
The highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Clothianidin (max(RQ) = 157), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Where analysed, metabolites exhibited the highest detection rates (for example, Metazachlor sulfonic acid was detected in 84\% of all samples were it was analysed (n = 3038, see also Supplemental Figure~S9).
Glyphosate was the compound with the highest detection rates (41\%, n = 3557 samples), followed by Boscalid (23\%, n = 9886) and Isoproturon (22\%, n = 19112). 
However, only the latter showed RQ exceedances (Figure~\ref{fig:fig6}).
In 45.9\% of samples more than one compound was quantified, with a maximum of 54 different compounds in one sample (Supplemental Figure~S8).

### 2016-10-12 supplement/Supplemental_Materials.tex
fixes to abstract and precipitation
restricted analyses to <100km2, revised ms accordingly
reanalysed with only data for < 100km2

%% -------------------------------------------------------------------------
\chapter{Catchment size - stream width relationships}
We studied the relationship between catchment size based on three datasets containing this informations:
Data delivered by the federal state Thuringia, \citet{vos_organic_2015} and \citet{fernandez_effects_2015} (both from Rhineland-Palatinate).
We fitted to each dataset separately and to the combined dataset a power-function.
The resulting models are shown in Figure \ref{fig:size_width}.

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{width_size}
	\caption[Relationship between catchment size and stream width.]{Relationship between catchment size and stream width. A power function has been fitted to each dataset separately and the combined dataset (black line and equation). The gray rectangle marks the estimated with for a catchment size of 25km\textsuperscript{2}.}
	\label{fig:size_width}
\end{figure}

@@

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{coefs}
	\caption[Graphical representation of coefficients from table \ref{tab:var_model_coef}.]{Graphical representation of coefficients from table \ref{tab:var_model_coef}. Top row: Effect of precipitation at day before sampling and at day of sampling. Bottom row: estimates for the four Quarters. Each dot represent one compound. Coefficients where the CI encompasses zero are shown in gray colour. Coefficients are shown on the link scale (log for $\mu$ and logit for $\nu$).}
	\label{fig:coefs}
\end{figure}

>>>>

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.6\textwidth]{silhouette}
	\caption[Average silhouette width for different cluster sizes.]{Average silhouette width for different cluster sizes. Two clusters showed maximum silhouette width.}
	\label{fig:varclus}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{coefs}
	\caption[Graphical representation of coefficients from table \ref{tab:var_model_coef}.]{Graphical representation of coefficients from table \ref{tab:var_model_coef}. Top row: Effect of precipitation at day before sampling and at day of sampling. Bottom row: estimates for the four Quarters. Each dot represent one compound (in the order described in table \ref{tab:var_model}). Coefficients where the CI encompasses zero are shown in gray colour. Coefficients are shown on the link scale (log for $\mu$ and logit for $\nu$).}
	\label{fig:coefs}
\end{figure}
